\tcode{Generalisation of nth_element to a range of nths}

### 2021-12-15 draft_lundberg_nth_elements.tex
Started Draft. Near-verbatim included (shortened down) version of performance study, with reference for details

\newenvironment{MyQuote}{%
    \begin{myQuoteEnumerate}[resume=*,series=MyQuoteSeries]%
    \color{blue}%
    \item \begin{quote}%
}{%
    \end{quote}%
    \end{myQuoteEnumerate}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%
%\iffalse
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{lipsum}
\makeatletter
\renewcommand{\linenumberfont}{\fontsize{12pt}{3pt}\normalfont\smaller\color{gray}}
\makeatother
%\usepackage{lineno}
%\linenumbers
%\fi
%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.2}
%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\renewcommand\tableofcontents{%
    \@starttoc{toc}%
}
\makeatother

@@

\begin{tabular}{ll}
Document \#:& P2375 D1 \\
Date: &2021-05-14 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

Fixed typo. (23021-12-06)

@@

\textbf{5} 
Complexity: For the overloads with no ExecutionPolicy, linear on average.
For the overloads with an ExecutionPolicy, O(N) applications of the predicate, and O(N log N) swaps, where N = last - first. 
\added{For overloads taking a range [nths_first,nths_last) but no ExecutionPolicy the complexity is approximately \mbox{O(N log m)} where m is the number of unique elements in [nths_first,nths_last).
For overloads taking a range [nths_first,nths_last) and an ExecutionPolicy the complexity is \mbox{O(N log N)}.}

@@

@@

>>>>

\newenvironment{MyQuote}{%
    \begin{myQuoteEnumerate}[resume=*,series=MyQuoteSeries]%
    \color{blue}%
    \item \begin{quote}%
}{%
    \end{quote}%
    \end{myQuoteEnumerate}%
}%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
%\iffalse
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{lipsum}
\makeatletter
\renewcommand{\linenumberfont}{\fontsize{12pt}{3pt}\normalfont\smaller\color{gray}}
\makeatother
%\usepackage{lineno}
%\linenumbers
%\fi
%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.2}
%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\renewcommand\tableofcontents{%
    \@starttoc{toc}%
}
\makeatother

@@

\begin{tabular}{ll}
Document \#:& D2375R1 W.I.P. 2021-12-15\\
Date: &2021-05-14 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

Fixed typo. (2021-12-06)\\
Summarized performance study, with references. (2021-12-15)\\

@@

\textbf{5} 
Complexity: For the overloads with no ExecutionPolicy, linear on average.
For the overloads with an ExecutionPolicy, O(N) applications of the predicate, and O(N log N) swaps, where N = last - first. 
\added{For overloads taking a range [nths_first,nths_last) but no ExecutionPolicy the complexity is approximately \mbox{O(N log m + M)} where m is the number of unique elements in [nths_first,nths_last) where M = nths_last - nths_first.
For overloads taking a range [nths_first,nths_last) and an ExecutionPolicy the complexity is \mbox{O(N log N + M)}.}

@@

\section{Reference implementation and practical Performance}

A reference implementation is found at \cite{p2375src}. A performance study is found at \cite{p2375RefImpl} and summarized here.

\subsubsection*{Time vs \tcode{std::sort} for equidistant partitioning points}

\begin{center}
\includegraphics[width=0.6\textwidth]{plotting/images/partition_points_equidistant_speed_for_n_vs_sort.png}
\end{center}

The above image shows the execution speed of \tcode{nth_element} compared to \tcode{std::sort} for various vector sizes $N$.
The data to sort here consists of randomly shuffled integers, and the range-of-nths version of \tcode{nth_element} was given different numbers $m$ of evenly spaced partitioning points in the vector. Dotted lines show lines $k_i \cdot log(N)$ fitted to pass though each curve at $N=3e6$.

The image shows that the speed curves approximately follow the expected $log(N)$ shape, with different factors for different $m$.

The following table shows a few speedup factors (rough read from the plot) for a number of unique partitioning points $m$ and vector sizes $N$.

%With $m = 1$ partitioning points the speed of \tcode{nth_elements} is about $10$ times that of \tcode{std::sort} at $N=3e7$.
%
%With $m = 5$ partitioning points the speed of \tcode{nth_elements} is about $5$ times that of \tcode{std::sort} at $N=3e7$.
%
%With $m = 500$ partitioning points the speed of \tcode{nth_elements} is about $2$ times that of \tcode{std::sort} at $N=3e7$.
%
%With $m = 50$ partitioning points the speed of \tcode{nth_elements} is about $2$ times that of \tcode{std::sort} at $N=3e5$.
%
%With $m = 5$ partitioning points the speed of \tcode{nth_elements} is about $2$ times that of \tcode{std::sort} at $N=3e3$.

\begin{center}
\begin{tabular}{r|c|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
  \textbf{m} & \textbf{speedup} & \textbf{at N}\\
  \hline 
1 & 10 & $N=3e7$\\
5 & 5 & $N=3e7$\\
500 & 2 & $N=3e7$\\
\hline
10 & 3 & $N=3e5$\\
50 & 2 & $N=3e5$\\
\hline
5 & 2 & $N=3e3$\\
\end{tabular}
\end{center}

Some care is required when interpreting the plot at low N: For example at the curve $m=1000$. $N=300$, the actual number of partitioning points are saturated to $N$, that is $300$ unique partitioning points. Still, one can say, that as $N$ decreases and approaches $m$ the performance is worse than \tcode{std::sort}, but the speed in these tests are still about $85\%$ of \tcode{std::sort} even for $m \sim N$. The benefit of \tcode{nth_element} over \tcode{sort} grows with $N$ as $log N$. Further slices and ways to plot same performance data is found at \cite{p2375RefImpl}. It also shows the study repeated for uniformly random (unique) partition points with very similar conclusion, but slightly better performance.

%\includegraphics[width=0.6\textwidth]{plotting/images/partition_points_equidistant_time_for_n_div.png}

%The above image shows a different view of the same data. The execution times for \tcode{std::sort} for various vector sizes $N$ and for \tcode{nth_element} with different numbers $m$ of evenly spaced partitioning points in the vector. Arbitrary normalization.

%The complexity of \tcode{nth_element} is $O(N log m)$. Therefore a thin line $k_i  N$ is also drawn for each curve $i$ where $k_i$ is set to match %at $N=3e6$. For \tcode{std::sort}, a dotted line $s × N log N$ is drawn instead. (Note how similar $N log N$ and $N$ looks in log-log.)

%\includegraphics[width=0.6\textwidth]{plotting/images/partition_points_equidistant_time_for_n_div.png}
%
%This is the same data as shown in the previous figure except all data points are divided by the $s × N log N$ curve described above.
%In other words, the image represents the execution time compared to the expected $N log N$ of \tcode{std::sort} at $N=3e6$. As expected, the \tcode{std::sort} %line approaches $1.0$ for large $N$, while the relative execution time for \tcode{nth_element} decreases as $1/log N$ for large N.
%
%\includegraphics[width=0.6\textwidth]{plotting/images/partition_points_equidistant_speed_for_m_vs_sort.png}
%
%Another view of the same data, but shows the speedup as function of $m$ for different vector sizes $N$.
%It's interesting that even for quite small $N=3e4$ there are speedups of $2$ to $6$ times for $m=1$ to $m=20$.

@@

\bibitem[p2375RefImpl]{p2375RefImpl}
Reference implementation of range-of-nths nth_element\\
\url{https://github.com/jmlundberg/nth_element_material}

\bibitem[p2375src]{p2375src}
Document source for \emph{this} proposal and reference implementation\\
\url{https://github.com/jmlundberg/p2375}
